---
title: "Class 12 (Online)"
author: 'Farnam (PID: A17628539)'
date: "2024-02-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1. proportion of G/G in a population

Downloaded a CSV file from ensemble <
https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39840102-39950103;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100

```

## Section 2 and 3 would be done in Galaxy Portal

Now let's dig into this further:

## One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. 

How many samples do we have?

```{r}
expr <- read.table("Last section.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```

Let's make a boxplot:

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) + 
geom_boxplot(notch = TRUE)

```









